The SMT Competition 2015--2018

### 2019-08-27 2018/report/SMT-COMP.tex
problem -> benchmark; more work on Scoring

\paragraph{Benchmarks with Unknown Status.}
Another paradigm shift concerns the treatment of benchmarks with
unknown status.  Because of the difficulty to judge solver responses
for such benchmarks, SMT Competitions until 2015 only used benchmarks
whose status was known (i.e., that had been solved previously by
existing SMT solvers).  In 2015, \num{30717} non-incremental benchmarks
from the SMT Library were not used for the competition because their
status was unknown.  However, this approach was criticized for
imitating existing solvers rather than truly innovating.  In~2016,
the SMT Library contained \num{29724} non-incremental benchmarks with
unknown status; solver performance on these was evaluated but reported
separately.  Since 2017, all non-incremental benchmarks are eligible
for the competition's \maintrack regardless of status, and they
directly affect the competition results.  Solver responses for such
benchmarks are assumed to be correct, unless there is a disagreement
between two or more solvers that are otherwise (i.e., on problems with
known status) sound.  Benchmarks with disagreements are removed from
the competition results.  This conservative approach was chosen over a
voting system as the latter is unreliable (especially in divisions
with few solvers), could lead to a solver being punished wrongly, and
could be gamed by entering multiple copies of a solver into the
competition.  Moreover, the size and time frame of the competition
demand an automated resolution.  Disagreements are reported to the
solver developers involved, who typically perform a manual
investigation at their leisure, and in all cases have led to bugs in
solvers being identified and fixed.  In 2016, there were disagreements
on 79 benchmarks; in 2017, only one benchmark was removed; while in
2018, there were no disagreements between otherwise sound solvers.

@@

  \caption{Number of benchmarks used in the \maintrack (by division). Numbers in parentheses indicate the (rounded) percentage of problems with different statuses, which are (sat/unsat) in 2015 and 2016, and (sat/unsat/unknown) in 2017 and 2018.}

@@

@@

The \emph{raw benchmark score} of a solver is a quadruple $\langle
e, n, w, c\rangle$, with
\begin{itemize}[noitemsep]
  \item \makebox[5em][l]{$e\geq 0$} number of erroneous results (usually~$e = 0$)
  \item \makebox[5em][l]{$0 \leq n \leq N$} number of correct results (resp.~\emph{reduction} for the \ucoretrack)
  \item \makebox[5em][l]{$w \in [0,T]$} wall-clock time in seconds (real-valued)
  \item \makebox[5em][l]{$c \in [0, 4T]$} CPU time in seconds (real-valued)
\end{itemize}
\noindent
For the \maintrack and \apptrack, $e$ is the number of returned statuses that disagree with the given expected status; $e \in \{0,1\}$ for the \maintrack.  (As described above, disagreements between solvers on benchmarks with unknown status lead to the benchmark being disregarded.)  For the \ucoretrack, $e$ includes, in addition, the number of returned unsat cores that are ill-formed or are not, in fact, unsatisfiable (as validated by a selection of other solvers selected by organizers).

@@

\subsection{Competition-Wide Ranking}
\label{sec:scoring:floc}

@@

It is interesting to note that the non-competing solvers MathSAT and
Z3 often appear as the best performing solver.  We view this as
confirmation of the industrial significance of the competition
results.  Both solvers are industrial-strength tools, actively
maintained and tuned for applications.  The reasons for non-entry into
the competition are orthogonal to their suitability for the
competition.  Indeed, Z3 has been entered into the competition
competitively in the past, and one can assume that it was trained on
the current problems.  The fact that industrial-strength solvers
perform well in the competition supports the notion that the
benchmarks in the competition are industrially relevant.

@@

Table~\ref{table:competition-wide-ranking} shows the best three
solvers according to the competition-wide ranking for 2015--2018.
Also shown is Z3, which ranked in high position each year but was
running hors concours.  Sequential and parallel performance are
reported separately.
%
Note that scores are not directly comparable between different years:
the set of benchmarks changes from one year to the other, and also the
formula to compute the score changed (see below).  Moreover, readers
should keep in mind that this ranking---like any method that maps the
performance of solvers for a broad range of logics to a
one-dimensional value---is inherently limited.  We refer to the
individual division rankings (Section~\ref{sec:division-rankings}) for
more detailed information about each solver.

@@

The 2018 edition of the competition was again part of the FLoC Olympic
Games, which sponsored seven medals.  The competition organizers
awarded the following solvers:
\begin{itemize}
\item The best three \maintrack solvers according to the
  competition-wide ranking (CVC4, Yices2, and SMTInterpol; see
  Table~\ref{table:competition-wide-ranking}),
\item the solvers that won the QF\_BV division in the \maintrack
  (Boolector and MinkeyRink; see Table~\ref{tab:winners-main}),
\item the solver that won the most competitive divisions in the
  \apptrack (Yices2),
\item the solver that won the most competitive divisions in the
  \ucoretrack (Yices2).
\end{itemize}

@@

For the \emph{\maintrack}, we determine the VBS as the best wall-clock
performance per (correctly solved) instance with the time limit of 1200s used
in the competitions 2017 and 2018. Figure~\ref{fig:progress-all} visualizes VBS
performance on common benchmarks from 2015--2018 as cactus plot.
Figure~\ref{fig:progress} shows VBS performance on common benchmarks over pairs
of consecutive years from 2014 to 2018.  Both plots show the wall-clock solving
time per instance over all instances, sorted by solving time.
%
Between 2014 (\num{67426} benchmarks) and 2015 (\num{154238} benchmarks),
the number of common benchmarks is \num{67070};
between 2015 and 2016 (\num{154424} benchmarks) it is \num{154238};
between 2016 and 2017 (\num{238758} benchmarks) it is \num{145236};
and between 2017 and 2018 (\num{333241} benchmarks) it is \num{226429}.
The number of solved instances on these sets of common benchmarks
increased from 2014 to 2015 by 0.1\%, from 2016 to 2017 by 3.9\%,
and from 2017 to 2018 by 0.3\%.
Interestingly, it decreased from 2015 to 2016 by 3.5\%.
This is mainly due to the fact that, in 2015, more than \num{5600} benchmarks of
division QF\_FP where solved by one of the two participating configurations of
Z3 that no one else could solve, and in~2016 Z3 did not participate in this
division.

For the \emph{\apptrack}, we determine the VBS as the best performance
in terms of number of correct \texttt{check-sat} answers within the time limit
of 2400s (the time limit of the \apptrack in the competitions
2015--2018).
The total number of eligible \texttt{check-sat} queries (i.e., all queries with
status sat or unsat up to the first query with status unknown) on
common benchmarks from 2015--2018 is \num{30370517}.
Figure~\ref{fig:progress-all-app} visualizes VBS performance on these
benchmarks as a bar plot.
In 2015, already 99.95\% of the eligible queries were answered correctly,
which naturally does not leave a lot of room for improvement.
In 2016, performance decreased by \num{185625} correct answers (0.61\%),
which can be entirely attributed to a performance regression
of Z3 on two benchmarks in the QF\_NIA division.

For the \emph{\ucoretrack}, we determine the VBS as the best performance in
terms of \emph{reduction}, i.e., the difference in the number of assertions
of the input formula and the unsat core, within the time limit
of 2400s (the time limit of the \ucoretrack in the competitions
2016--2018).
Figure~\ref{fig:progress-all-app} visualizes VBS performance on the common
benchmarks (\num{91455} total) from 2016--2018 as a bar plot.
For \num{32377} of these benchmarks, extracting an unsat core is trivial since
they contain only one single \texttt{assert} command.
They are thus also not considered in the bar plot, since their maximum
possible reduction is trivially 0.

@@

In many cases, an SMT application produces either mainly satisfiable or mainly
unsatisfiable queries to the SMT solver.  A typical example is bounded model checking,
where a query is only satisfiable if a property of the model does not hold.  As
a consequence, some solvers may implement specialized techniques and
optimizations for the more frequent (the expected) case.
%
Therefore, it is interesting to ask whether the winners of divisions of a track
might have changed if benchmarks in that division had been restricted to one of
these results.  Tables~\ref{tab:results:sat}-\ref{tab:results:unsat} show the
changes in division winners if only satisfiable or unsatisfiable results are
given. This table is in reference to Table~\ref{tab:winners-main} and only
lists a solver if the result has changed. A non-competing solver is listed by
itself if it was not the overall best performing solver of that division but
now outperforms the winning competitive solver, and the winning competing
solver did not change.
It is interesting to note that we see some new solvers winning divisions that did not win any division in the main results. For example, Alt-Ergo solves the most satisfiable problems in AUFLIA and ALIA in 2018, and SMTInterpol performs very well on unsatisfiable problems in QF\_LIA.

We only consider the \maintrack here, which contains a mixture of problems that
are satisfiable and unsatisfiable. A percentage of satisfiable, unsatisfiable
and unknown status benchmarks per division and year is given in
Table~\ref{table:benchmarks-main-track}.  For the \apptrack, this analysis is
less meaningful since a benchmark usually contains a multitude of both
satisfiable and unsatisfiable incremental queries
(which may depend on each other).
For the \ucoretrack, only known unsatisfiable benchmarks from the \maintrack
are used, which renders this analysis useless for this track.

>>>>

\paragraph{Benchmarks with Unknown Status.}
Another paradigm shift concerns the treatment of benchmarks with
unknown status.  Because of the difficulty to judge solver responses
for such benchmarks, SMT Competitions until 2015 only used benchmarks
whose status was known (i.e., that had been solved previously by
existing SMT solvers).  In 2015, \num{30717} non-incremental benchmarks
from the SMT Library were not used for the competition because their
status was unknown.  However, this approach was criticized for
imitating existing solvers rather than truly innovating.  In~2016,
the SMT Library contained \num{29724} non-incremental benchmarks with
unknown status; solver performance on these was evaluated but reported
separately.  Since 2017, all non-incremental benchmarks are eligible
for the competition's \maintrack regardless of status, and they
directly affect the competition results.  Solver responses for such
benchmarks are assumed to be correct, unless there is a disagreement
between two or more solvers that are otherwise (i.e., on benchmarks with
known status) sound.  Benchmarks with disagreements are removed from
the competition results.  This conservative approach was chosen over a
voting system as the latter is unreliable (especially in divisions
with few solvers), could lead to a solver being punished wrongly, and
could be gamed by entering multiple copies of a solver into the
competition.  Moreover, the size and time frame of the competition
demand an automated resolution.  Disagreements are reported to the
solver developers involved, who typically perform a manual
investigation at their leisure, and in all cases have led to bugs in
solvers being identified and fixed.  In 2016, there were disagreements
on 79 benchmarks; in 2017, only one benchmark was removed; while in
2018, there were no disagreements between otherwise sound solvers.

@@

  \caption{Number of benchmarks used in the \maintrack (by division). Numbers in parentheses indicate the (rounded) percentage of benchmarks with different statuses, which are (sat/unsat) in 2015 and 2016, and (sat/unsat/unknown) in 2017 and 2018.}

@@

The \emph{raw benchmark score} of a solver is thus a quadruple
$\langle e, n, w, c\rangle$, with
\begin{itemize}[noitemsep]
  \item \makebox[5em][l]{$e\geq 0$} number of erroneous responses (usually~$e = 0$),
  \item \makebox[5em][l]{$0 \leq n \leq N$} number of correct responses (resp.~\emph{reduction} for the \ucoretrack),
  \item \makebox[5em][l]{$w \in [0,T]$} wall-clock time in seconds (real-valued),
  \item \makebox[5em][l]{$c \in [0, 4T]$} CPU time in seconds (real-valued).
\end{itemize}
%
For the \maintrack and \apptrack, $e$ is the number of returned
statuses that disagree with the given expected status; $e \in \{0,1\}$
for the \maintrack.  (As described in Section~\ref{sec:benchmarks},
disagreements between otherwise sound solvers on benchmarks with
unknown status lead to the benchmark being disregarded.)  For the
\ucoretrack, $e$ includes, in addition, the number of returned unsat
cores that are ill-formed or are not, in fact, unsatisfiable (as
validated by a selection of other solvers selected by organizers).

@@

@@

\subsection{Competition-Wide Ranking}
\label{sec:scoring-floc}

@@

It is interesting to note that the non-competing solvers MathSAT and
Z3 often appear as the best performing solver.  We view this as
confirmation of the industrial significance of the competition
results.  Both solvers are industrial-strength tools, actively
maintained and tuned for applications.  The reasons for non-entry into
the competition are orthogonal to their suitability for the
competition.  Indeed, Z3 has been entered into the competition
competitively in the past, and one can assume that it was trained on
the current benchmarks.  The fact that industrial-strength solvers
perform well in the competition supports the notion that the
benchmarks in the competition are industrially relevant.

@@

Table~\ref{table:competition-wide-ranking} shows the best three
\maintrack solvers according to the competition-wide ranking for
2015--2018.  Also shown is Z3, which ranked in high position each year
but was running hors concours.  Sequential and parallel performance
are reported separately.
%
Note that scores are not directly comparable between different years:
the set of benchmarks changes from one year to the other, and also the
formula to compute the score changed (see
Section~\ref{sec:scoring-floc}).  Moreover, readers should keep in
mind that this ranking---like any method that maps the performance of
solvers for a broad range of logics to a one-dimensional value---is
inherently limited.  We refer to the individual division rankings
(Section~\ref{sec:division-rankings}) for more detailed information
about each solver.

@@

The 2018 edition of the competition was part of the FLoC Olympic
Games, which sponsored seven medals.  The competition organizers
awarded the following solvers:
\begin{itemize}
\item The best three \maintrack solvers according to the
  competition-wide ranking (CVC4, Yices2, and SMTInterpol; see
  Table~\ref{table:competition-wide-ranking}),
\item the solvers that won the QF\_BV division in the \maintrack
  (Boolector and MinkeyRink; see Table~\ref{tab:winners-main}),
\item the solver that won the most competitive divisions in the
  \apptrack (Yices2; see Table~\ref{tab:winners-application}),
\item the solver that won the most competitive divisions in the
  \ucoretrack (Yices2; see Table~\ref{tab:winners-unsat-core}).
\end{itemize}

@@

For the \maintrack, we determine the VBS as the best wall-clock
performance per (correctly solved) instance with the time limit of \SI{1200}{s} used
in the competitions 2017 and 2018. Figure~\ref{fig:progress-all} visualizes VBS
performance on common benchmarks from 2015--2018 as cactus plot.
Figure~\ref{fig:progress} shows VBS performance on common benchmarks over pairs
of consecutive years from 2014 to 2018.  Both plots show the wall-clock solving
time per instance over all instances, sorted by solving time.
%
Between 2014 (\num{67426} benchmarks) and~2015 (\num{154238} benchmarks),
the number of common benchmarks is \num{67070};
between 2015 and 2016 (\num{154424} benchmarks) it is \num{154238};
between 2016 and 2017 (\num{238758} benchmarks) it is \num{145236};
and between 2017 and 2018 (\num{333241} benchmarks) it is \num{226429}.
The number of solved instances on these sets of common benchmarks
increased from 2014 to 2015 by 0.1\%, from 2016 to 2017 by 3.9\%,
and from 2017 to 2018 by 0.3\%.
Interestingly, it decreased from 2015 to 2016 by 3.5\%.
This is mainly due to the fact that, in 2015, more than \num{5600} benchmarks of
division QF\_FP where solved by one of the two participating configurations of
Z3 that no one else could solve, and in~2016 Z3 did not participate in this
division.

For the \apptrack, we determine the VBS as the best performance
in terms of number of correct \texttt{check-sat} answers within the time limit
of \SI{2400}{s} (the time limit of the \apptrack in the competitions
2015--2018).
The total number of eligible \texttt{check-sat} queries (i.e., all queries with
status sat or unsat up to the first query with status unknown) on
common benchmarks from 2015--2018 is \num{30370517}.
Figure~\ref{fig:progress-all-app} visualizes VBS performance on these
benchmarks as a bar plot.
In 2015, already 99.95\% of the eligible queries were answered correctly,
which naturally does not leave a lot of room for improvement.
In 2016, performance decreased by \num{185625} correct answers (0.61\%),
which can be entirely attributed to a performance regression
of Z3 on two benchmarks in the QF\_NIA division.

For the \ucoretrack, we determine the VBS as the best performance in
terms of reduction, i.e., the difference in the number of assertions
of the input formula and the unsat core, within the time limit
of \SI{2400}{s} (the time limit of the \ucoretrack in the competitions
2016--2018).
Figure~\ref{fig:progress-all-app} visualizes VBS performance on the common
benchmarks (\num{91455} total) from 2016--2018 as a bar plot.
For \num{32377} of these benchmarks, extracting an unsat core is trivial since
they contain only one single \texttt{assert} command.
They are thus not considered in the bar plot, since their maximum
possible reduction is trivially 0.

@@

In many cases, an SMT application produces either mainly satisfiable or mainly
unsatisfiable queries to the SMT solver.  A typical example is bounded model checking,
where a query is only satisfiable if a property of the model does not hold.  As
a consequence, some solvers may implement specialized techniques and
optimizations for the more frequent (the expected) case.
%
Therefore, it is interesting to ask whether the winners of divisions of a track
might have changed if benchmarks in that division had been restricted to one of
these results.  Tables~\ref{tab:results:sat}-\ref{tab:results:unsat} show the
changes in division winners if only satisfiable or unsatisfiable results are
given. This table is in reference to Table~\ref{tab:winners-main} and only
lists a solver if the result has changed. A non-competing solver is listed by
itself if it was not the overall best performing solver of that division but
now outperforms the winning competitive solver, and the winning competing
solver did not change.
It is interesting to note that we see some new solvers winning divisions that did not win any division in the main results. For example, Alt-Ergo solves the most satisfiable benchmarks in AUFLIA and ALIA in 2018, and SMTInterpol performs very well on unsatisfiable benchmarks in QF\_LIA.

We only consider the \maintrack here, which contains a mixture of benchmarks that
are satisfiable and unsatisfiable. A percentage of satisfiable, unsatisfiable
and unknown status benchmarks per division and year is given in
Table~\ref{table:benchmarks-main-track}.  For the \apptrack, this analysis is
less meaningful since a benchmark usually contains a multitude of both
satisfiable and unsatisfiable incremental queries
(which may depend on each other).
For the \ucoretrack, only known unsatisfiable benchmarks from the \maintrack
are used, which renders this analysis useless for this track.
